---
title: "Red Wine Analysis"
author: "Harshal Kulkarni"
date: "October 10 2018"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

library(ggplot2)
library(GGally)
library(RColorBrewer)
library(reshape)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv("wineQualityReds.csv")
```

# Introduction
The analysis that follows is based on the following dataset:
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

# Exploratoring the dataset

## An overview of the data
The following variables are included in the Red Wine dataset:

```{r echo=FALSE}
names(wine)
```

And the variables are of type:

```{r echo=FALSE}
str(wine)
```

So it seems the X variable is just an index for each observation in the dataset, and the remaining variables are quantified using numerical data. The Quality variable is provided as integers.

Let's look at the variability in the numerical data:

```{r echo=FALSE}
wine <- subset(wine, select = -c(X))
summary(wine)
```


We can visualize the variability of each variable by plotting each variable using a boxplot:

```{r echo=FALSE}
#we use the reshape library to melt the wide data and then visualize the data distribution for each variable using boxplots. 

meltData <- melt(wine)
ggplot(meltData, aes(factor(variable), value))+ 
  geom_boxplot() + facet_wrap(~variable, scale="free")
```


The upper and lower whiskers extend to the highest and lowest point whithin 1.5 times the inter-quartile range

We can also plot histograms for each variable to help us understand the distribution of each variable:

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(meltData, aes(x = value)) + 
    facet_wrap(~variable,scales = "free_x") + 
    geom_histogram()
```

Most of the variables are normally distributed.  Alcohol, sulphates and total sulfur dioxide seem to have lognormal distributions. The outliers for the residual sugar and chlorides variables make it difficult to see the distribution. Let's exclude the 95th percentile for residual sugar and chlorides and re-plot their histograms:

```{r echo=FALSE, warning=FALSE, message=FALSE}
#calculating the quantile
#quantile(wine$residual.sugar, c(0.99))
 p1 <- ggplot(wine, aes(x=residual.sugar)) + 
  geom_histogram() + 
  xlim(0, quantile(wine$residual.sugar, c(0.95)))    
p2 <- ggplot(wine, aes(x=chlorides)) + 
  geom_histogram() + 
  xlim(0, quantile(wine$chlorides, c(0.95)))
grid.arrange(p1, p2, ncol=2)
```

The distributions for residual sugar and chlorides also look normal after we exclude the outliers.

Here are the summary statistics for residual sugar:

```{r echo=FALSE}
summary(wine$residual.sugar)
```

And chlorides:

```{r echo=FALSE}
summary(wine$chlorides)
```

What we are most interested in is the quality rating and what variables influence the quality rating. Let's take a look at the red wine quality ratings using a histogram.

```{r echo=FALSE}
ggplot(wine, aes(x=quality)) + 
  geom_histogram(binwidth=1)
summary(wine$quality)
```

Majority of red wines have a quality ranking of 5 and 6.

The alcohol content can be another important consideration when we are purchasing wine:

```{r echo=FALSE}
ggplot(wine, aes(x=alcohol)) + 
  geom_density()
summary(wine$alcohol)
```

It looks like the alcohol content of the wine in the dataset follows a lognormal distribution with a high peak at the lower end of the alcohol scale.


# Bivariate Analysis

We can quickly visualize the relationship between each pair of variables and find their pearson product-moment correlation. The names along the x and y axis of the plot matrix below are as follows:

```{r echo=FALSE}
names(wine)
```

```{r echo=FALSE}
#ggpairs(wine, columns = c(2:13), columnLabels = c("f.acidity", "v.acidity", "c.acid", "res.sugar", "chlorides", "free.SO2", "total.SO2", "density", "pH", "sulphates", "alcohol","quality")) + theme(axis.title.y = element_text(size = 8))

ggscatmat(wine) + theme_grey(base_size=7)
```

The top 4 correlation coefficients with quality are:
* alchol:quality = 0.48
* sulphates:quality = 0.25
* citric.acid:quality = 0.23
* fixed.acidity:quality = 0.12

So as we saw earlier, alcohol content has a high correlation with red wine quality.  Other important attributes correlated with red wine quality include sulphates, citric acid and fixed acidity.

The biggest negative corrlation coefficients with quality are:
* volatile.acidity:quality = -0.39
* total.sulfur.dioxide:quality = -0.19
* density:quality = -0.17
* chlorides:quality = -0.13

So we see that volatile acids are negatively correlated with red wine quality.  Tota sulfur dioxide, density and chlorides are also negatively correlated with quality. 

Variables with the highest (positive or negative) correlation include:
* fixed.acidity:citirc.acid = 0.67
* fixed.acidity:density = 0.67
* free.sulfur.dioxide:total.sulfur.dioxide = 0.67
* alcohol:quality = 0.48
* density:alcohol = -0.50
* citric.acid:pH = -0.54
* volatile.acidity:citirc.acid = -0.55
* fixed.acidity:pH = -0.68

Let's look at a few of these relationships in a bit more detail.  Density and alcohol:

```{r echo=FALSE}
ggplot(wine, aes(density, alcohol)) + 
  geom_jitter(alpha=0.2) + 
  stat_smooth(method = "lm", formula = y ~ x)
```

We see that density tends to increase with decreasing alcohol content.

Fixed acidity and pH:

```{r echo=FALSE}
ggplot(wine, aes(fixed.acidity, pH)) + 
  geom_jitter(alpha=0.2) + 
  stat_smooth(method = "lm", formula = y ~ x)
```

We also see fixed acidity increase as pH decreases.

Fixed acidity and density:

```{r echo=FALSE}
ggplot(wine, aes(fixed.acidity, density)) + 
  geom_point(alpha=0.2) +  
  stat_smooth(method = "lm", formula = y ~ x)
```

Fixed acidity increases with density.

Now let's look at the alcohol content by red wine quality using a density plot function:

```{r echo=FALSE}
#I tried running the plot below but there was an error, likely because quality is an int, not a factor so I made quality a factor
ggplot(wine, aes(x=alcohol, fill=factor(quality))) + 
  geom_density(alpha=0.2)
```

It seems the red wines with a higher alcohol content tend to have a higher quality rating...what a surprise!  The main anomoly to this trend appears to be red wines having a quality ranking of 5.

Here are the summary statistics for alcohol content at each quality level:

```{r echo=FALSE}
by(wine$alcohol, factor(wine$quality), summary)
```

```{r echo=FALSE}
ggplot(wine, aes(x=sulphates, fill=factor(quality))) + 
  geom_density(alpha=0.2)
```

It seems sulphate content is quite important for red wine quality, particularly for the highest quality levels including quality 7 and 8.

And here are the summary statistics for sulphates at each quality level:

```{r echo=FALSE}
tapply(wine$sulphates, factor(wine$quality), summary)
```



Now let's use function "factor" as a factor with default order:

```{r}
wine$quality <- factor(wine$quality)
ggplot(data = wine, aes(x = quality, y = alcohol)) +
  geom_jitter(alpha = 1/10) +
  geom_boxplot(alpha = 1/10, color = 'blue') +
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red') +
  labs(x = 'quality levels',
       y = 'alcohol (% by volume)',
       title = 'Boxplot of alcohol along each quality level')
  
```

As seen before majority of red wines have 5 and 6 as quality rating. The box plot confirms this further. 


## Multivariate Plots


Now lets visualize relationship between alochol and density with the help of Scatterplot:

```{r echo=FALSE, warning=FALSE, message=FALSE}
wine$quality <- factor(wine$quality)
ggplot(data = wine, aes(x = density, y = alcohol, color = quality)) +
  geom_jitter() +
  coord_cartesian(xlim = c(0.985, 1.0)) +
  scale_color_brewer(type = 'div', palette = "PuOr") +
  theme_dark() +
  labs(x = 'density (mg/l)',
       y = 'alcohol (% by volume)',
       title = 'Scatterplot between density and alcohol with colored quality levels')
```


It seems that more the alcohol (% by vloume) better the quality of wine


Now let's visualize the relationship between sulphates, volatile.acidity, alcohol and quality:


```{r echo=FALSE}
ggplot(wine, aes(x=volatile.acidity, y=sulphates, 
                 color=factor(quality))) + 
  geom_point() + 
  scale_color_brewer(type = "seq", palette="1")
```

It seems the higher quality red wines tend to be concentrated in the left of the plot. This tends to be where the higher alcohol content (larger dots) are concentrated as well.


Let's try summarizing quality using a contour plot of alcohol and sulphate content:

```{r echo=FALSE}
ggplot(wine, aes(x = sulphates, y = alcohol, 
                 colour=factor(wine$quality))) + 
  geom_density2d(bins=2) + 
  scale_color_brewer() + 
  geom_point(color="black", alpha=0.1) 
```

This shows that higher quality red wines are generally located near the upper right of the scatter plot (darker contour lines) wheras lower quality red wines are generally located in the bottom right.


Let's make a similar plot but this time quality will be visualized using density plots along the x and y axis and color :

```{r echo=FALSE}
empty <- ggplot()+geom_point(aes(1,1), colour="white") +
     theme(                              
       plot.background = element_blank(), 
       panel.grid.major = element_blank(), 
       panel.grid.minor = element_blank(), 
       panel.border = element_blank(), 
       panel.background = element_blank(),
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
       axis.text.x = element_blank(),
       axis.text.y = element_blank(),
       axis.ticks = element_blank()
     )
 main <- ggplot(wine, aes(x = sulphates, y = alcohol, 
                 color=factor(wine$quality))) + 
  geom_point() +
  theme(legend.position=c(1,1),legend.justification=c(1,1))
 top <- ggplot(wine, aes(x = sulphates, 
                 fill=factor(wine$quality))) + 
  geom_density(alpha=0.5) + 
  theme(legend.position = "none")
  
right <- ggplot(wine, aes(x = alcohol, 
                 fill=factor(wine$quality))) + 
  coord_flip() + 
  geom_density(alpha=0.5) + 
  theme(legend.position = "none")
  
grid.arrange(top, empty, main, right, ncol=2, nrow=2, widths=c(4, 1), heights=c(1, 4))
```

Again, this clearly illustrates that higher quality wines are found near the top right of the plot.

#Final Plots & Summary
Now let's summarize the main findings with a few refined plots.

The strongest correlation coefficient was found between alcohol and quality.  Now let's look at the alcohol content by red wine quality using a density plot function:

```{r echo=FALSE}
#I tried running the plot below but there was an error, likely because quality is an int, not a factor so I made quality a factor
colours = brewer.pal(name="RdBu", n=nlevels(factor(wine$quality)))
names(colours) = rev(levels(factor(wine$quality)))
 ggplot(wine, aes(x=alcohol, fill=factor(quality))) + 
  geom_density(alpha=0.2) + 
  scale_fill_manual(values=colours) + 
  labs(title="Density plot of alcohol content 
       by red wine quality ratings", 
       x="alcohol (% by volume)") + 
  theme(plot.title = element_text(size = 14))
```

Clearly we see that the density plots for higher quality red wines (as indicated by the red plots) are right shifted, meaning they have a comparatively high alcohol content, compared to the lower quality red wines.  However, the main anomoly to this trend appears to be red wines having a quality ranking of 5.

Here are the summary statistics for alcohol content at each quality level:

```{r echo=FALSE}
by(wine$alcohol, wine$quality, summary)
```

Previously Sulphates were also found to be correlated with red wine quality (R^2= 0.25) while volatile acid had a negative correlation (R^2=-0.39).  We can visualize the relationships betwen these two variables, along with alcohol content and red wine quality using a scatter plot:

```{r echo=FALSE}
wine$alcohol.levels <- cut(wine$alcohol, breaks=c(0,10,12,14,max(wine$alcohol)),
                           labels=c("Low (< 10)", "Medium (10 to 12)", "High (12 to 14)", "Very High (>14)"))
colours = brewer.pal(name="RdBu", n=nlevels(factor(wine$quality)))
names(colours) = rev(levels(factor(wine$quality)))

ggplot(wine, aes(x=volatile.acidity, y=sulphates, 
                 fill=factor(quality))) + 
  geom_point(shape=21) + 
  scale_fill_manual(values=colours) + 
  labs(title="Influences Sulphates, and Volatile Acidity on Red Wine Quality\
       Faceted by Alcohol Content,", 
       x = (expression(Volatile~Acidity~(g/dm^{3}))),
       y = (expression(Sulphates~(Potassium~Sulphate~-~g/dm^{3})))) + 
  theme(plot.title = element_text(size = 14)) +
   facet_wrap(~alcohol.levels)
```

We see a clear trend where higher quality red wines(red dots) are concentrated in upper right of the figure. 

And here is a summary of red wine alcohol content by quality rating:

```{r echo=FALSE}
by(wine$alcohol, wine$quality, summary)
```

By sulphate content:

```{r echo=FALSE}
by(wine$sulphates, wine$quality, summary)
```

And by volatile.acidity

```{r echo=FALSE}
by(wine$volatile.acidity, wine$quality, summary)
```

We can also visualize the relationship between alcohol content and sulphates by combining scatter plots with density plots:

```{r echo=FALSE}
empty <- ggplot()+geom_point(aes(1,1), colour="white") +
     theme(                              
       plot.background = element_blank(), 
       panel.grid.major = element_blank(), 
       panel.grid.minor = element_blank(), 
       panel.border = element_blank(), 
       panel.background = element_blank(),
       axis.title.x = element_blank(),
       axis.title.y = element_blank(),
       axis.text.x = element_blank(),
       axis.text.y = element_blank(),
       axis.ticks = element_blank()
     )
 main <- ggplot(wine, aes(x = sulphates, y = alcohol, 
                 fill=factor(wine$quality))) + 
  geom_point(shape=21) +
  scale_fill_manual(values=colours) +
  theme(legend.position=c(1,1),legend.justification=c(1,1)) + 
  labs(fill="Quality", 
       title="Influence of alcohol and sulphate 
       content on red wine quality", 
       y="alcohol content (% by volume)", 
       x= "sulphates (potassium sulphate - g / dm3)")
  
 top <- ggplot(wine, aes(x = sulphates, 
                 fill=factor(wine$quality))) + 
  geom_density(alpha=0.5) + 
  scale_fill_manual(values=colours) +
  theme(legend.position = "none")
  
right <- ggplot(wine, aes(x = alcohol, 
                 fill=factor(wine$quality))) + 
  coord_flip() + 
  geom_density(alpha=0.5) + 
  scale_fill_manual(values=colours) +
  theme(legend.position = "none")
  
grid.arrange(top, empty, main, right, ncol=2, nrow=2, widths=c(4, 1), heights=c(1, 4))
```

By combining a scatter plot with density plots of the x- and y-axis variables it is easier to see that a trend from low to high quality moving from the bottom left of the scatter plot that is blue to the top left of the scatter plot which is red, but we can density plots on each axis help illustrate the rather distinct ranges that are occupied for different quality wines for both alcohol content and sulphate content. 

#Reflection

 The above analysis considered the relationship of a number of red wine attributes with the quality rankings of different wines.  Melting the dataframe and using facet grids was really helpful for visualizing the distribution of each of the parameters with the use of boxplots and histograms. Most of the parameters were found to be normally distributed while citirc acid, free sulfur dioxide and total sulfur dioxide and alcohol had more of a lognormal distribution.

 GGally provides a concise summary of the pairwise relationships in the dataset although the scatter plots can easily become overplotted. Formatting the axis was also an issue because of the overlapping of the axis labels with the tick mark labels.  I finally settled on the `ggscatmat()` plot which moved the axis labels to the top and right rather than overlapping the tick labels as `ggpairs()` does.  
 
 Using the insights from correlation coefficients provided by the paired plots, it was interesting exploring quality using density plots with a different color for each quality.  Once I had this plotted it was interesting to build up the multivariate scatter plots to visualize the relationship of different variables with quality by also varying the point size, using density plots on the x and y axis, and also using density plots.
 
Next step would be to develop a statistical model to predict red wine quality based on the data in this dataset.
